---
title: "A field guide to base R"
format: html
---

## Exercises 27.2.4 {.unnumbered}

### Question 1

> Create functions that take a vector as input and return:

**Solution:**

> The elements at even-numbered positions.

```{r, eval=FALSE}
even_numbered <- function(x){
  c = 1:length(x)
  x[c %% 2 == 0]
}
```

> Every element except the last value.

```{r, eval=FALSE}
except_last <- function(x){
  x[1:length(x)-1]
}
```

> Only even values (and no missing values).

```{r}
even_non_missing <- function(x){
  x[x%%2 == 0 & !is.na(x)]
}
```

### Question 2

> Why is `x[-which(x > 0)]` not the same as `x[x <= 0]`? Read the documentation for `which()` and do some experiments to figure it out.

**Solution:**

> In this specific example where the vector contains only negative values `x <- c(-1, -2)`. `x[x <= 0]` will return `c(-1, -2)` and in the other hand `x[-which(x > 0)]` will output numeric(0). The first uses logical subsetting and the second integer subsetting. 

## Exercises 27.3.4 {.unnumbered}

### Question 1

> What happens when you use `[[` with a positive integer that’s bigger than the length of the vector? What happens when you subset with a name that doesn’t exist?

**Solution:**

> In both cases it raises an error except for a list it returns NULL.

### Question 2 

> What would `pepper[[1]][1]` be? What about `pepper[[1]][[1]]`?

**Solution:**

> I guess `pepper[[1]][1]` would extract a smaller packet containing only the first grain of pepper. Therefore, `pepper[[1]][[1]]` would subsequently extract a single grain of pepper.
