---
title: "Communication"
---

## Exercises 11.2.1 {.unnumbered}

### Question 1

> Create one plot on the fuel economy data with customized title, subtitle, caption, x, y, and color labels.

**Solution:**

```{r}
library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)

ggplot(mpg, 
       aes(x = class, y = hwy))+
  geom_boxplot()+
  coord_flip()+
  labs(
    x = "Type of car",
    y = "highway miles per gallon",
    title = "Distribution of hwy across different car type",
    subtitle = "Big cars definitely need more energy (fuel) to move their massive weight and push through\nthe air, that's why suvs and puckups have lower highway fuel efficiency",
    caption = "Data from fueleconomy.gov"
  )
```

### Question 2

> Recreate the following plot using the fuel economy data. Note that both the colors and shapes of points vary by type of drive train.

**Solution:**

```{r}
ggplot(mpg, 
       aes(x = cty, y = hwy, color = drv, shape = drv))+
  geom_point()+
  labs(color = "Type of drive train", 
       shape = "Type of drive train")
```

### Question 3

> Take an exploratory graphic that you’ve created in the last month, and add informative titles to make it easier for others to understand.

**Solution:**

```{r}
library(lvplot)
ggplot(mpg, 
       aes(x = class, y = hwy))+
  geom_lv()+
  labs(
    x = "Type of car",
    y = "highway miles per gallon",
    title = "Distribution of hwy across different car type",
    subtitle = "Big cars tend to consume more energy (fuel) to move their massive weight, support itself and push\nthrough the air: that's why suvs and puckups have lower highway fuel efficiency",
    caption = "Data from fueleconomy.gov"
  )+
  annotate(
    "label", 
    x = "suv",       # Centers the text above the 'suv' column
    y = 38,          # Y-coordinate high enough to avoid overlapping the data
    label = "These two heavy vehicle types\nshow the lowest efficiency.", 
    color = "darkred", 
    fontface = "bold",
    hjust = 0.8      # Shifts the text slightly to the left to sit between pickup and suv
  )+
  annotate(
    "segment", 
    x = "suv", xend = "suv", 
    y = 35, yend = 25, 
    arrow = arrow(length = unit(0.3, "cm")), 
    color = "darkred",
    linewidth = 1
  )+
  annotate(
    "curve",         # Using a curve here to reach over the 'subcompact' column
    x = "suv", xend = "pickup", 
    y = 36, yend = 25, 
    curvature = 0.3,
    arrow = arrow(length = unit(0.3, "cm")), 
    color = "darkred",
    linewidth = 1
  )
```

## Exercises 11.3.1 {.unnumbered}

### Question 1 & 2

> Use geom_text() with infinite positions to place text at the four corners of the plot.
>
> Use annotate() to add a point geom in the middle of your last plot without having to create a tibble. Customize the shape, size, or color of the point.

**Solution:**

> Make sure to read the code to identify what's each of its components does/adds to the plot. I refer to each component by a comment to figure out its utility.

```{r}
# 1. Create a mini-dataset for the four corners
corner_labels <- data.frame(
  x_coord = c(-Inf, -Inf, Inf, Inf),
  y_coord = c(-Inf, Inf, -Inf, Inf),
  text_label = c("Bottom-Left", "Top-Left", "Bottom-Right", "Top-Right")
)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.3) +
  annotate(geom = "point", x = 4.5, y = 30, # Question 2 - we add the point in the middle
           size = 5, 
           color = "steelblue")+
  annotate(
    geom = "label", x = 3, y = 37.2,
    label = "Larger engine sizes tend to\nhave lower fuel economy",
    hjust = "left", color = "red"
  ) +
  annotate(
    geom = "segment",
    x = 3, y = 35, xend = 5, yend = 25, color = "red",
    arrow = arrow(type = "closed", angle = 20)
  )+
  geom_text(          # Question 1
    data = corner_labels, 
    aes(x = x_coord, y = y_coord, label = text_label),
    hjust = "inward", # Pushes text away from the left/right edges
    vjust = "inward", # Pushes text away from the top/bottom edges
    color = "steelblue",
    fontface = "bold",
    size = 4.2
  )
```

### Question 3

> How do labels with geom_text() interact with faceting? How can you add a label to a single facet? How can you put a different label in each facet? (Hint: Think about the dataset that is being passed to geom_text() )

**Solution:**

> Whenever you are working with facets and annotations: Always pass a separate `data.frame` to your annotation layer (like `geom_text`), and ensure that `data.frame` contains the faceting variable. This gives you total control over exactly what text appears in which box.

> Below, you find a code that illustrate this mechanism

```{r}
#| warning: false
#| layout-ncol: 1
#| fig-width: 10

# Create a multi-row dataset with a unique label for each facet
multi_label_data <- data.frame(
  drv = c("4", "f", "r"),  # The three facets
  hwy = c(40, 40, 40),     # Y coordinates for each label
  displ = c(5, 5, 5),      # X coordinates for each label
  label = c(
    "4WD vehicles", 
    "Front-wheel drive", 
    "Rear-wheel drive"
  )
)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ drv) +
  geom_text(
    data = multi_label_data, 
    aes(label = label),
    color = "blue",
    fontface = "bold"
  )
```

### Question 4

> What arguments to geom_label() control the appearance of the background box?

**Solution:**

> fill, color and alpha. Respectively, it controls the background color of the box and Controls the color of the box's outer border. Alpha controls the transparency of the box.

### Question 5

> What are the four arguments to arrow()? How do they work? Create a series of plots that demonstrate the most important options.

**Solution:**

> The four arguments of `arrow()` are : angle, length, ends & type. Feel free to look at the default values and tweak them to change it's appearance.

## Exercises 11.4.6 {.unnumbered}

### Question 1

> Why doesn’t the following code override the default scale?

**Solution:**

> To override the default blue gradient inside the hexagons, you must `use scale_fill_gradient()` instead.

```{r}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)

ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_fill_gradient(low = "white", high = "red") +
  coord_fixed()
```

### Question 2

> What is the first argument to every scale? How does it compare to labs()?

**Solution:**

> The first argument to every single `scale_* function` is name. This argument controls the text label that appears as the axis title or the legend title for that specific scale. labs & scale do the exact same thing for axis and legend titles.

### Question 3

> Change the display of the presidential terms by:

> - Combining the two variants that customize colors and x axis breaks.
> - Improving the display of the y axis.
> - Labelling each term with the name of the president.
> - Adding informative plot labels.
> - Placing breaks every 4 years.

**Solution:**

```{r}
library(ggplot2)
library(dplyr)

# a
presidential |> 
  mutate(id = 33 + row_number()) |> 
  ggplot(aes(x = start, y = id, color = party)) + 
  geom_point() + 
  geom_segment(aes(xend = end, yend = id)) + 
  # Variant 1: Customizing the x-axis breaks and labels
  scale_x_date(
    name = NULL, 
    breaks = presidential$start, 
    date_labels = "'%y"
  ) +
  # Variant 2: Customizing the colors based on the political party
  scale_color_manual(
    values = c("Republican" = "red", "Democratic" = "blue")
  )

# b 

library(forcats) # For working with categorical data (factors)

presidential |> 
  mutate(
    id = 33 + row_number(),
    # Create a clean label combining the number and the name (e.g., "34. Eisenhower")
    pres_label = paste0(id, ". ", name),
    # Force the new labels to be ordered by their start date, not alphabetically
    pres_label = fct_reorder(pres_label, start)
  ) |> 
  
  # Map y to our newly ordered label instead of 'id'
  ggplot(aes(x = start, y = pres_label, color = party)) + 
  geom_point() + 
  geom_segment(aes(xend = end, yend = pres_label)) + 
  
  scale_x_date(
    name = NULL, 
    breaks = presidential$start, 
    date_labels = "'%y"
  ) +
  scale_color_manual(
    values = c("Republican" = "red", "Democratic" = "blue")
  ) +
  
  # Remove the y-axis title entirely, as the names are self-explanatory
  labs(y = NULL)

# c 

presidential |> 
  mutate(id = 33 + row_number()) |> 
  ggplot(aes(x = start, y = id, color = party)) + 
  geom_point() + 
  geom_segment(aes(xend = end, yend = id)) + 
  
  # Add the text labels
  geom_text(
    aes(label = name), 
    hjust = 0,         # Left-align the text with the start point
    nudge_y = 0.3,     # Nudge the text slightly UP so it doesn't overlap the line
    size = 3.5,        # Adjust text size
    show.legend = FALSE # Prevents the text from messing up your legend formatting
  ) +
  
  scale_x_date(
    name = NULL, 
    breaks = presidential$start, 
    date_labels = "'%y"
  ) +
  scale_color_manual(
    values = c("Republican" = "red", "Democratic" = "blue")
  )

# d

presidential |> 
  mutate(id = 33 + row_number()) |> 
  ggplot(aes(x = start, y = id, color = party)) + 
  geom_point() + 
  geom_segment(aes(xend = end, yend = id)) + 
  
  geom_text(
    aes(label = name), 
    hjust = 0,         
    nudge_y = 0.3,     
    size = 3.5,        
    show.legend = FALSE 
  ) +
  
  scale_x_date(
    name = NULL, 
    breaks = presidential$start, 
    date_labels = "'%y"
  ) +
  scale_color_manual(
    values = c("Republican" = "red", "Democratic" = "blue")
  ) +
  # Add all the informative plot labels here
  labs(
    title = "U.S. Presidential Terms",
    subtitle = "Timeline of presidencies from Eisenhower to Obama",
    caption = "Data source: ggplot2::presidential",
    color = "Political Party", 
    y = NULL 
  )

# e 

presidential |> 
  mutate(id = 33 + row_number()) |> 
  ggplot(aes(x = start, y = id, color = party)) + 
  geom_point() + 
  geom_segment(aes(xend = end, yend = id)) + 
  
  geom_text(
    aes(label = name), 
    hjust = 0,         
    nudge_y = 0.3,     
    size = 3.5,        
    show.legend = FALSE 
  ) +
  
  # Updated x-axis scale:
  scale_x_date(
    name = NULL, 
    date_breaks = "4 years", # Forces a tick mark every 4 years
    date_labels = "%Y"       # Uses the full 4-digit year (e.g., 1960)
  ) +
  
  scale_color_manual(
    values = c("Republican" = "red", "Democratic" = "blue")
  ) +
  
  labs(
    title = "U.S. Presidential Terms",
    subtitle = "Timeline of presidencies from Eisenhower to Obama",
    caption = "Data source: ggplot2::presidential",
    color = "Political Party", 
    y = NULL 
  )
```
### Question 4 

> First, create the following plot. Then, modify the code using `override.aes` to make the legend easier to see.

**Solution:**

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 1/20) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))
```

## Exercises 11.5.1 {.unnumbered}

### Question 1

> Pick a theme offered by the ggthemes package and apply it to the last plot you made.

**Solution:**

> I am choosing them_bw() function to adopt a white background with grid lines.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 1/20) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  theme_bw()
```

### Question 2 

> Make the axis labels of your plot blue and bolded.

**Solution:**

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), alpha = 1/20) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))+
  labs(x = "Carat Weight", y = "Price (USD)")+
  theme(
    axis.title = element_text(color = "blue", face = "bold")
  )
```
## Exercises 11.6.1  {.unnumbered}

### Question 1

> What happens if you omit the parentheses in the following plot layout. Can you explain why this happens?

```{r}
p1 <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 1")
p2 <- ggplot(mpg, aes(x = drv, y = hwy)) + 
  geom_boxplot() + 
  labs(title = "Plot 2")
p3 <- ggplot(mpg, aes(x = cty, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 3")

p1 | p2 / p3
```

**Solution:**

> Instead of having Plot 1 and Plot 2 side-by-side on the top row with Plot 3 spanning the entire bottom row, you will get a two-column layout.

>In R, apparently,  the division operator (/) has a higher mathematical precedence than the logical OR operator (|). Therefore, the expression p1 | p2 / p3 is evaluated as p1 | (p2 / p3).

### Question 2 

> Using the three plots from the previous exercise, recreate the following patchwork.

**Solution:**

```{r}
library(ggplot2)
library(patchwork)

# 1. Define the three individual plots
p1 <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 1")

p2 <- ggplot(mpg, aes(x = drv, y = hwy)) + 
  geom_boxplot() + 
  labs(title = "Plot 2")

p3 <- ggplot(mpg, aes(x = cty, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 3")

# 2. Combine and tag the plots using patchwork
patchwork_layout <- p1 / (p2 | p3)

patchwork_layout + 
  plot_annotation(
    tag_levels = "A", 
    tag_suffix = ":",
    tag_prefix = "Fig. "
  )
```

