---
title: "Databases"
format: html
---

## Exercises 21.5.10  {.unnumbered}

### Question 1 

> What is distinct() translated to? How about head()?

**Solution:**

> As you can notice below, `distinct()` simply becomes `DISTINCT` and since `head()` shows the first 6 rows of our dataset, it gets translated to `LIMIT 6`

```{r}
library(DBI)
library(dbplyr)
library(tidyverse)

con <- DBI::dbConnect(duckdb::duckdb())
dbplyr::copy_nycflights13(con)

flights <- tbl(con, "flights")
planes <- tbl(con, "planes")
```

```{r}
flights |> 
  distinct(dest) |> 
  show_query()
```
```{r}
planes |> 
  head() |> 
  show_query()
```
### Question 2 

> Explain what each of the following SQL queries do and try recreate them using dbplyr.

```{r, eval=FALSE}
# 1
SELECT * 
FROM flights
WHERE dep_delay < arr_delay

# 2
SELECT *, distance / (air_time / 60) AS speed
FROM flights
```

**Solution:**

> It is pretty straghtforward to explain what each of these SQL queries does. The first one keeps rows based on a condition and the second one create a new column named speed based on the right components that forms its expression.


```{r}
# 1
flights |> 
  filter(dep_delay < arr_delay) |> 
  show_query()
```

```{r}
# 2 
flights |> 
  mutate( speed = distance / (air_time / 60)) |> 
  show_query()
```



