---
title: "Data import"
---

## Exercises 7.2.4 {.unnumbered}

### Question 1

> What function would you use to read a file where fields were separated with `“|”`?

**Solution:**

> Since `read_delim()` reads in files with any delimiter, and attempts to automatically guess the delimiter if you don’t specify it. I would try it first without specifying the `“|”` delimiter & inspect the result or doing it from the very begining and directly set the given delimiter.

### Question 2

> Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?

**Solution:**

> The `read_csv()` and `read_tsv()` functions both have the `col_types` argument in common, which allows you to specify the data types for each column.

### Question 3

> What are the most important arguments to read_fwf()?

**Solution:**

> The `read_fwf()` function from the `readr` package (part of the `tidyverse`) is used to import fixed-width files—data formats where columns are aligned by specific character widths or positions rather than separated by delimiters like commas or tabs.

> The two most important arguments you need to specify are:

1.  `file`: The path to the file, a connection, or a raw string containing the data.
2.  `col_positions` : This is the most critical argument specific to fixed-width files. It tells R where **each column starts and ends.**

### Question 4

> Sometimes strings in a CSV file contain commas. To prevent them from causing problems, they need to be surrounded by a quoting character, like `"` or `'`. By default, `read_csv()` assumes that the quoting character will be ". To read the following text into a data frame, what argument to `read_csv()` do you need to specify?

**Solution:**

> In this example, we encounter a data frame where the first column `x` contains numerical values and the second one contains strings as values ('a,b' as a whole represents a value of the variable y). To do so, we specify the argument `col_types` as `"dc"` : `double` for the first variable and `character` for the second one. The code is below.

```{r}
library(tidyverse)
read_csv(I("x,y\n1,'a,b'"), col_types = "dc")
```

### Question 5

> Identify what is wrong with each of the following inline CSV files. What happens when you run the code?

**Solution:**

```{r, eval=FALSE}
# 1 |
read_csv("a,b\n1,2,3\n4,5,6")

# 2 | 
read_csv("a,b,c\n1,2\n1,2,3,4")

# 3 | 
read_csv("a,b\n\"1")

# 4 | 
read_csv("a,b\n1,2\na,b")

# 5 | 
read_csv("a;b\n1;3")
```

> The first code creates two variables `a` & `b` along with rows of three cell values. theoretically, it is quite impossible to execute but read_csv finds always a way to represent our data in unexpected ways as we will see in the examples below. Try to execute the code to see how read_csv interprets those inputs.

> The second code creates " variables `a`, `b` & `c`. Introduce a row with two cell values and then a second row with 4 cell values. Another example of miss-using those inputs. But an opportunity of exploring how `read_csv` handles inconsistent data.

> The third code ideally should create two columns and fill rows with values for each one of them. But here, we notice a strange typo that follows the new line character. Surprisingly, read_csv found a way to alter column types and fill the rows with related values. Try to execute the code to see how read_csv interprets those inputs.

> The fourth code : Here things gets more interesting. As we start to read, from left to right, we might expect the first column to be numerical (`double`) but the second rows introduce strings along with those numbers. read_csv altered the column types and converted numbers to character to keep uniformity along each column.

> The fifth code introduces data separated with semi-colon instead of being delimited with coma since we are using `read_csv` (that's the human read but the machine has another way to interprets things) : here `read_csv` considers "a;b" as a column non-syntactic name (that's why it wrapped it up between back-ticks) along with a single value as a character which is '1;3'.

### Question 6

> Practice referring to non-syntactic names in the following data frame by:

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```


a.  Extracting the variable called `1`
b.  Plotting a scatter plot of `1` vs. `2`
c.  Creating a new column called `3`, which is `2` divided by `1`
d.  Renaming the columns to `one`, `two`, and `three`

**Solution:**

> Even though digits are allowed to be used in syntactic names, it should not start with it. For example `data.2026` is a valid syntactic name but `2026.data` is not.

a.  Extracting the variable called `1` 

```{r, eval=FALSE}
annoying |> 
  select(`1`)
```

b.  Plotting a scatter plot of `1` vs. `2`

```{r, eval=FALSE}
library(ggplot2)

annoying |> 
  ggplot(aes(x=`1`, y=`2`)) +
  geom_point()
```

c.  Creating a new column called `3`, which is `2` divided by `1`

```{r, eval=FALSE}
annoying |> 
  mutate(`3` = `2`/`1`)
```

d.  Renaming the columns to `one`, `two`, and `three`

```{r, eval=FALSE}
annoying |> 
  mutate(`3` = `2`/`1`) |> 
  rename(one = `1`, two = `2`, three = `3`)
```



















