---
title: "Dates and times"
format: html
---

## Exercises 17.2.5    {.unnumbered}

### Question 1 

> What happens if you parse a string that contains invalid dates?

```{r, eval=FALSE}
ymd(c("2010-10-10", "bananas"))
```

**Solution:**

> It simply fails to parse/transform it into a valid date.

### Question 2 

> What does the tzone argument to today() do? Why is it important?

**Solution:**

> `tzone` is a character vector specifying which time zone you would like the current time in. tzone defaults to your computer's system timezone.

### Question 3 

> For each of the following date-times, show how you’d parse it using a readr column specification and a lubridate function.

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"
```

**Solution:**

```{r, eval=FALSE}
#d1
mdy(d1)

#d2
ymd(d2)

#d3
dmy(d3)

#d4
mdy(d4)

#d5
mdy(d5)

#t2
# November day 15, 2010, 12 PM (noon)
csv <- "
  date
  11:15:10.12 PM
"
read_csv(csv, col_types = cols(date = col_datetime("%m:%d:%y%.%I%.%p")))

#t1
# Simply 5 PM (17h) & 5 minutes
csv_2 <- "
  time
  1705
"
read_csv(csv_2, col_types = cols(time = col_time("%H%M")))

```

