---
title: "Factors"
format: html
---

## Exercises 16.3.1   {.unnumbered}

### Question 1 

> Explore the distribution of `rincome` (reported income). What makes the default bar chart hard to understand? How could you improve the plot?

**Solution:**

> As you may notice if we ever want to plot categories on the x axis, given their strings length, they become stacked one beside each other which leads to unpleasing visibility. To make it look better, we simply the axes.

```{r}
#| warning: false
#| layout-ncol: 2
#| fig-width: 5

library(ggplot2)
library(tidyverse)

# first plot: categories on the x-axis
gss_cat |> 
  count(rincome) |> 
  ggplot(aes(x = rincome, y = n )) + 
  geom_col() + 
  labs(title = "Distribution of Reported Income", x = "Income Category", y = "Count")

# second plot: categories on the y-axis
gss_cat |> 
  count(rincome) |> 
  ggplot(aes(x = n, y = rincome)) + 
  geom_col() + 
  labs(title = "Distribution of Reported Income", x = "Count", y = "Income Category" )
```

### Question 2 

> What is the most common `relig` in this survey? What’s the most common `partyid`?

**Solution:**

```{r}
#| warning: false
#| layout-ncol: 2
#| fig-width: 5

# the most common religions
gss_cat |> 
  count(relig) |> 
  arrange(desc(n)) |> 
  head(5)

# the most common party affiliation
gss_cat |> 
  count(partyid) |> 
  arrange(desc(n)) |> 
  head(5)
```
### Question 3 

> Which relig does denom (denomination) apply to? How can you find out with a table? How can you find out with a visualization?

**Solution:**

> It is pretty straghtforward from the table. When inspecting the plotn we confirm our findings in the table. almost every religion defaults to the category 'not applicable' except `protestant`, `christian` ...

```{r}
# Count all combinations of religion and denomination
gss_cat |> 
  count(relig, denom) |> 
  # Remove the rows where denomination was marked "Not applicable"
  filter(denom != "Not applicable") |> 
  # Look at which religions remain
  distinct(relig)

gss_cat |> 
  count(relig, denom) |> 
  ggplot(aes(x = relig, y = denom, size = n)) +
  geom_point(color = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
by_age <- gss_cat |>
  filter(!is.na(age)) |>
  count(age, marital) |>
  group_by(age) |>
  mutate(
    prop = n / sum(n)
  )
```

