---
title: "Functions"
format: html
---

## Exercises 25.2.5 {.unnumbered}

### Question 1

> Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need?

```{r, eval=FALSE}
mean(is.na(x))
mean(is.na(y))
mean(is.na(z))

x / sum(x, na.rm = TRUE)
y / sum(y, na.rm = TRUE)
z / sum(z, na.rm = TRUE)

round(x / sum(x, na.rm = TRUE) * 100, 1)
round(y / sum(y, na.rm = TRUE) * 100, 1)
round(z / sum(z, na.rm = TRUE) * 100, 1)
```

**Solution:**

```{r, eval=FALSE}
# 1st function : Percentage of non available data
# for example : mean(is.na(c(1, NA))) is equal to 50% which means half of our data is missing as NA
# is.na() returns a logical vector with TRUE when the data is missing and FALSE otherwise.
# The mean function recognizes TRUE as 1 and FALSE as 0. It adds all the zeros and ones and divide by the
# length of our vector.

Percentage_na <- function(x){
  mean(is.na(x))
}

# The second function rescales the vector to have a range from 0 to 1 and obviously return an NA value 
# when the actual data is missing

rescale02 <- function(x){
  x / sum(x, na.rm = TRUE)
}

# The third function is built on top of rescale02, it actually just change the way we read the rescaled values
# from the second function. It gives the percentage of each value relatively to the sum of recorded data and 
# keep only one decimal number after the comma. So when defining this function we may give ourself a bit of 
# freedom to play around with the fractional part.

percentage_rescale <- function(x, nbr_decimal){
  round(rescale02(x)*100, nbr_decimal)
}

```

### Question 2

> In the second variant of `rescale01()`, infinite values are left unchanged. Can you rewrite `rescale01()` so that `-Inf` is mapped to 0, and `Inf` is mapped to 1?

**Solution:**

```{r, eval=FALSE}
library(tidyverse)
library(nycflights13)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  case_when(
    x == -Inf ~ 0,
    x ==  Inf ~ 1,
    .default = (x - rng[1]) / (rng[2] - rng[1])
  )
}

#  x <- c(NA, 1:3, -1:1/0)
#> rescale01(x)
#  [1]  NA 0.0 0.5 1.0 0.0 NaN 1.0
```

### Question 3

> Given a vector of birthdates, write a function to compute the age in years.

**Solution:**

```{r, eval=FALSE}
get_age <- function(birthdays) {
  birthdays <- as_date(birthday)
  (birthdays %--% today()) %/% years(1)
}
```

### Question 4

> Write your own functions to compute the variance and skewness of a numeric vector. You can look up the definitions on Wikipedia or elsewhere.

**Solution:**

> the goal of this exercise is to translate the raw mathematical formulas into R code using basic vector operations like `sum()`, `mean()`, and `length()`, explicitly avoiding `var()` or external packages.

```{r, eval=FALSE}
my_variance <- function(x) {
  n <- length(x)
  x_bar <- mean(x)
  sq_dev <- (x - x_bar)^2
  sum(sq_dev) / (n - 1)
}

my_skewness <- function(x) {
  n <- length(x)
  x_bar <- mean(x)
  m3 <- sum((x - x_bar)^3) / n
  m2 <- sum((x - x_bar)^2) / n
  m3 / (m2^(3/2))
}
```

### Question 5

> Write `both_na()`, a summary function that takes two vectors of the same length and returns the number of positions that have an NA in both vectors.

**Solution:**

```{r, eval=FALSE}
both_na <- function(x, y){
  if(length(x) == length(y)){
    sum(is.na(x)) + sum(is.na(y))
  }
  else{
    cat("Call again the function both_na and provide 2 vectors of the same length")
  }
}
```

### Question 6

> Read the documentation to figure out what the following functions do. Why are they useful even though they are so short?

```{r, eval=FALSE}
is_directory <- function(x) {
  file.info(x)$isdir
}
is_readable <- function(x) {
  file.access(x, 4) == 0
}
```

**Solution:**

> The `file.info()` function takes as input a vector of characters as files paths and return a data frame containing informations about those files. This data frame contain a logical column called `isdir` to indicate if the file is a directory or not. Therefore, `is_directory()` simply return given the path if the file is actually a directory or not.

> `file.access()` actually takes two arguments : a character vector containing file names and a second argument called mode whic is an integer specifying access mode. `is_readable()` is actually interessted in knowing wether the given file have a read permission or not and that's why it's called `is_readable()`. The output of `file.access()` is either 0 for a success or -1 for failure.

## Exercises 25.3.5 {.unnumbered}

### Question 1

> Using the datasets from nycflights13, write a function that:

**Solution:**

> Finds all flights that were cancelled (i.e. is.na(arr_time)) or delayed by more than an hour.

```{r, eval=FALSE}
filter_severe <- function(data){
  data |> 
    filter(is.na(arr_time) | dep_delay >= 60)
}
```

> Counts the number of cancelled flights and the number of flights delayed by more than an hour.

```{r, eval=FALSE}
summarize_severe <- function(data) {
  data |> 
    summarize(
      cancelled = sum(is.na(arr_time)),
      delayed = sum(dep_delay >= 60, na.rm = TRUE), 
      
    )
}
```

> Finds all flights that were cancelled or delayed by more than a user supplied number of hours:

```{r, eval=FALSE}
filter_severe <- function(data, hours){
  data |> 
    filter(is.na(arr_time) | dep_delay >= hours)
}
```

> Summarizes the weather to compute the minimum, mean, and maximum, of a user supplied variable:

```{r, eval=FALSE}
summarize_weather <- function(data, var){
  data |> 
    select(temp:visib) |> 
    summarize(minimum  = min({{var}}, na.rm = TRUE), 
              mean = mean({{var}}, na.rm = TRUE), 
              maximum = max({{var}}, na.rm = TRUE)
              )
}
```

> Converts the user supplied variable that uses clock time (e.g., dep_time, arr_time, etc.) into a decimal time (i.e. hours + (minutes / 60))

```{r, eval=FALSE}
standardize_time <- function(data, var){
  data |> 
    mutate({{ var }} := {{ var }}%/%100 + (({{ var }}%%100)/60) )
}
```

### Question 2

> For each of the following functions list all arguments that use tidy evaluation and describe whether they use data-masking or tidy-selection: `distinct()`, `count()`, `group_by()`, `rename_with()`, `slice_min()`, `slice_sample()`.

**Solution:**

- **`distinct()`**: The `...` argument uses **data-masking**. It evaluates the data values to find unique rows and allows you to compute new variables on the fly (e.g., `distinct(x * 2)`).

- **`count()`**: The `...` and `wt` arguments use **data-masking**. It calculates frequencies based on the data values of the provided columns or expressions.

- **`group_by()`**: The `...` argument uses **data-masking**. You can pass existing columns or create new grouping variables dynamically based on cell values (e.g., `group_by(total = a + b)`).

- **`rename_with()`**: The `.cols` argument uses **tidy-selection**. It dictates which columns to apply the renaming function to based entirely on their names or positions (like `starts_with("time")`), completely ignoring the data values inside them.

- **`slice_min()`**: The `order_by` argument uses **data-masking**. The function must look at the actual numerical values inside the specified column to rank them and find the minimums.

- **`slice_sample()`**: The `weight_by` argument uses **data-masking**. It evaluates the numeric values within a column to assign statistical sampling weights to each specific row.

### Question 3 

> Generalize the following function so that you can supply any number of variables to count.

```{r, eval=FALSE}
count_prop <- function(df, var, sort = FALSE) {
  df |>
    count({{ var }}, sort = sort) |>
    mutate(prop = n / sum(n))
}
```

**Solution:**

```{r, eval=FALSE}
count_prop <- function(df, ..., sort = FALSE) {
  df |>
    count(..., sort = sort) |>
    mutate(prop = n / sum(n))
}
```

## Exercises 25.4.4 {.unnumbered}

### Questions 1, 2, 3

> Draw a scatterplot given dataset and x and y variables. (*)

> Add a line of best fit (i.e. a linear model with no standard errors). (**)

> Add a title. (***)

**Solution:**

```{r, eval=FALSE}
library(ggplot2)
library(rlang)

scatter_plot <- function(df, var1, var2){
  
  label <- rlang::englue("A scatterplot of {{ var1 }} Vs {{ var2 }}")
  
  df |> 
    ggplot(aes(x = {{ var1 }}, y = {{ var2 }})) +
    geom_point() + #(*)
    geom_smooth(method = "loess", formula = y ~ x, color = "steelblue", se = FALSE) + #(**)
    labs(title = label) #(***)
}
```


## Exercises 25.5.1 {.unnumbered}

### Question 1 

> Read the source code for each of the following two functions, puzzle out what they do, and then brainstorm better names.

```{r, eval=FALSE}
f1 <- function(string, prefix) {
  str_sub(string, 1, str_length(prefix)) == prefix
}

f3 <- function(x, y) {
  rep(y, length.out = length(x))
}
```

**Solution:**

> To begin with, f1 takes two arguments and returns either `TRUE` or `FALSE`. Basically, It checks if the supplied string start with the given prefix. I would name it `is_prefix()`. Note that this function is case sensitive. Let's take this example below :

```{r, eval=FALSE}
hw <- "Hadley Wickham"

is_prefix <- function(string, prefix) {
  str_sub(string, 1, str_length(prefix)) == prefix
}

is_prefix(hw, "hadley") # hw starts with capital "H" and not "h"
 #> [1] FALSE
```

> f3 takes two vectors of values and return a vector of elements extracted from the second variable. What determines the length of the returned vector ? well, it comes from the length of the first vector. To put it simply : the returned vector contains elements from the second supplied vector with the length of the first vector. What's obvious is the following : if the length of the two vectors are equals the function simply return the second vector. If the length of x is smaller, it simply returns the first `length(x)` elements of the `y` vector. And finally if `length(x) > length(y)` we simply return the replicates in the same order of the elements of y until reaching the full length of `x`. 

> Personally, I would name it simply `replicate_length()`. I also would like to swap the variables to emphasize that we replicating the the first vector based on the length of the second one. 

```{r, eval=FALSE}
replicate_length <- function(y, x) {
  rep(y, length.out = length(x))
}
```


### Question 2 

> Take a function that you’ve written recently and spend 5 minutes brainstorming a better name for it and its arguments.

**Solution:**

> Let's go back to the plot function coded previously. It simply takes a data frame and plots a scatter plot of two of its variable (obviously numeric) against each other. 

```{r, eval=FALSE}
scatter_plot <- function(df, var1, var2){
  
  label <- rlang::englue("A scatterplot of {{ var1 }} Vs {{ var2 }}")
  
  df |> 
    ggplot(aes(x = {{ var1 }}, y = {{ var2 }})) +
    geom_point() + #(*)
    geom_smooth(method = "loess", formula = y ~ x, color = "steelblue", se = FALSE) + #(**)
    labs(title = label) #(***)
}
```

### Question 3 

> Make a case for why `norm_r()`, `norm_d()` etc. would be better than `rnorm()`, `dnorm()`. Make a case for the opposite. How could you make the names even clearer?

**Solution:**

> I would rename them as follows : 

- `norm_generate()` instead of `rnorm()`

- `norm_density()` instead of `dnorm()`