---
title: "Functions"
format: html
---

## Exercises 25.2.5 {.unnumbered}

### Question 1

> Practice turning the following code snippets into functions. Think about what each function does. What would you call it? How many arguments does it need?

```{r, eval=FALSE}
mean(is.na(x))
mean(is.na(y))
mean(is.na(z))

x / sum(x, na.rm = TRUE)
y / sum(y, na.rm = TRUE)
z / sum(z, na.rm = TRUE)

round(x / sum(x, na.rm = TRUE) * 100, 1)
round(y / sum(y, na.rm = TRUE) * 100, 1)
round(z / sum(z, na.rm = TRUE) * 100, 1)
```

**Solution:**

```{r}
# 1st function : Percentage of non available data
# for example : mean(is.na(c(1, NA))) is equal to 50% which means half of our data is missing as NA
# is.na() returns a logical vector with TRUE when the data is missing and FALSE otherwise.
# The mean function recognizes TRUE as 1 and FALSE as 0. It adds all the zeros and ones and divide by the
# length of our vector.

Percentage_na <- function(x){
  mean(is.na(x))
}

# The second function rescales the vector to have a range from 0 to 1 and obviously return an NA value 
# when the actual data is missing

rescale02 <- function(x){
  x / sum(x, na.rm = TRUE)
}

# The third function is built on top of rescale02, it actually just change the way we read the rescaled values
# from the second function. It gives the percentage of each value relatively to the sum of recorded data and 
# keep only one decimal number after the comma. So when defining this function we may give ourself a bit of 
# freedom to play around with the fractional part.

percentage_rescale <- function(x, nbr_decimal){
  round(rescale02(x)*100, nbr_decimal)
}

```

### Question 2

> In the second variant of `rescale01()`, infinite values are left unchanged. Can you rewrite `rescale01()` so that `-Inf` is mapped to 0, and `Inf` is mapped to 1?

**Solution:**

```{r, eval=FALSE}
library(tidyverse)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  case_when(
    x == -Inf ~ 0,
    x ==  Inf ~ 1,
    .default = (x - rng[1]) / (rng[2] - rng[1])
  )
}

#  x <- c(NA, 1:3, -1:1/0)
#> rescale01(x)
#  [1]  NA 0.0 0.5 1.0 0.0 NaN 1.0
```

### Question 3

> Given a vector of birthdates, write a function to compute the age in years.

**Solution:**

```{r}
get_age <- function(birthdays) {
  birthdays <- as_date(birthday)
  (birthdays %--% today()) %/% years(1)
}
```

### Question 4

> Write your own functions to compute the variance and skewness of a numeric vector. You can look up the definitions on Wikipedia or elsewhere.

**Solution:**

> the goal of this exercise is to translate the raw mathematical formulas into R code using basic vector operations like `sum()`, `mean()`, and `length()`, explicitly avoiding `var()` or external packages.

```{r}
my_variance <- function(x) {
  n <- length(x)
  x_bar <- mean(x)
  sq_dev <- (x - x_bar)^2
  sum(sq_dev) / (n - 1)
}

my_skewness <- function(x) {
  n <- length(x)
  x_bar <- mean(x)
  m3 <- sum((x - x_bar)^3) / n
  m2 <- sum((x - x_bar)^2) / n
  m3 / (m2^(3/2))
}
```

### Question 5

> Write `both_na()`, a summary function that takes two vectors of the same length and returns the number of positions that have an NA in both vectors.

**Solution:**

```{r}
both_na <- function(x, y){
  if(length(x) == length(y)){
    sum(is.na(x)) + sum(is.na(y))
  }
  else{
    cat("Call again the function both_na and provide 2 vectors of the same length")
  }
}
```

### Question 6

> Read the documentation to figure out what the following functions do. Why are they useful even though they are so short?

```{r, eval=FALSE}
is_directory <- function(x) {
  file.info(x)$isdir
}
is_readable <- function(x) {
  file.access(x, 4) == 0
}
```

**Solution:**

> The `file.info()` function takes as input a vector of characters as files paths and return a data frame containing informations about those files. This data frame contain a logical column called `isdir` to indicate if the file is a directory or not. Therefore, `is_directory()` simply return given the path if the file is actually a directory or not.

> `file.access()` actually takes two arguments : a character vector containing file names and a second argument called mode whic is an integer specifying access mode. `is_readable()` is actually interessted in knowing wether the given file have a read permission or not and that's why it's called `is_readable()`. The output of `file.access()` is either 0 for a success or -1 for failure.

## Exercises 25.3.5 {.unnumbered}

### Question 1

> Using the datasets from nycflights13, write a function that:

