---
title: "Iteration"
format: html
---

## Exercises 26.2.8 {.unnumbered}

### Question 1 

> Practice your `across()` skills by:

**Solution:**

> Computing the number of unique values in each column of `palmerpenguins::penguins`.

```{r}
library(tidyverse)

penguins |> 
  summarize(
    across(everything(), 
           function(x) n_distinct(x, na.rm = TRUE)
           )
    )
```

> Computing the mean of every column in mtcars.

```{r}
mtcars |> 
  summarize(
    across(everything(), 
           function(x) mean(x, na.rm = TRUE)
           )
  )
```

> Grouping diamonds by cut, clarity, and color then counting the number of observations and computing the mean of each numeric column.

```{r}
diamonds |> 
  group_by(cut, clarity, color) |> 
  summarize(across(where(is.numeric),
                   function(x) mean(x, na.rm = TRUE)
                   ), 
            n = n(), 
            .groups = "drop"
            )
```

### Question 2 

> What happens if you use a list of functions in `across()`, but don’t name them? How is the output named?

**Solution:**

> Let's take the data-frame given in Hadley book to illustrate what happens and for simplicity reasons. And let's compare when we use a list of functions and naming them vs where it's not the case. 

> You can identify such a difference in naming the new columns in the table below : concatenating the names with '_' as a separator versus simply adding numbers.

```{r}
df_date <- tibble(
  name = c("Amy", "Bob"),
  date = ymd(c("2009-08-03", "2010-01-16"))
)
```

```{r}
#| warning: false
#| layout-ncol: 2
#| fig-width: 5

# Using a named list of function
df_date |> 
  mutate(
    across(where(is.Date), list(year = year, month = month, day = mday))
    )

# Simply adding those functions
df_date |> 
  mutate(
    across(where(is.Date), 
           list(year, month, mday)
           )
    )
```

### Question 3 

> Adjust `expand_dates()` to automatically remove the date columns after they’ve been expanded. Do you need to embrace any arguments?

**Solution:**

> Since if we leave the result as it is, we kind of create a sorte of redundance. The question asks to to remove the columns from which we created the new ones. We can acheive this simply by adding a select statment as show below.

```{r, eval=FALSE}
expand_dates <- function(df) {
  df |> 
    mutate(
      across(where(is.Date), list(year = year, month = month, day = mday))
    ) |> 
    select(!where(is.Date))
}
```

### Question 4 

> Explain what each step of the pipeline in this function does. What special feature of where() are we taking advantage of?

```{r, eval=FALSE}
show_missing <- function(df, group_vars, summary_vars = everything()) {
  df |> 
    group_by(pick({{ group_vars }})) |> 
    summarize(
      across({{ summary_vars }}, \(x) sum(is.na(x))),
      .groups = "drop"
    ) |>
    select(where(\(x) any(x > 0)))
}
```

**Solution:**

> This function returns a data frame containing information about the number of missing values for each column after grouping by the supplied variables. The select function keep only columns where there is at least a missing value per grouping, so for example the column origin do not have a single missing value so it is dropped from the final data frame. 




